---
title: "Overview"
output: github_document
always_allow_html: true
date: "`r Sys.Date()`"
---


```{r, setup, echo=FALSE}
knitr::opts_chunk$set(
  comment = "#>",
  collapse = TRUE,
  warning = FALSE,
  message = FALSE,
  echo = FALSE,
  fig.width = 6,
  fig.asp = 0.618,
  out.width = "99%",
  fig.align = "center",
  dpi = 300
)
options(scipen = 999, digits = 2)
knitr::knit_hooks$set(
  inline = function(x) {
    if (is.numeric(x)) {
      return(prettyNum(x, big.mark = ","))
    } else{
      return(x)
    }
  }
)
```

```{r dependencies}
library(tidyverse, warn.conflicts = FALSE)
library(kableExtra)
```

```{r data, cache=TRUE}
library(hoaddata)

ta_oa_inst <- readr::read_csv(here::here("data", "ta_oa_inst.csv.gz"))
```

What are the number and proportion of open access articles in hybrid journals published under a transformative agreement?

```{r overview_data_prep}
ta_oa_growth <- ta_oa_inst |>
  filter(ta_active == TRUE, !is.na(cc), issn_l != "0027-8424", !cr_year %in%  c("2017", "2023")) |>
  group_by(cr_year) |>
  summarise(ta_oa = n_distinct(doi))
hybrid_oa_growth <- hoaddata::cc_articles |>
  filter(issn_l != "0027-8424", !cr_year %in%  c("2017", "2023")) |>
  group_by(cr_year) |>
  summarise(oa_all = n_distinct(doi))
ta_df <- inner_join(ta_oa_growth, hybrid_oa_growth) |>
  mutate(ta_oa_prop = ta_oa / oa_all) |>
  mutate(no_ta = oa_all - ta_oa, 
         no_ta_prop = 1 - ta_oa_prop) |>
  mutate(cr_year = as.factor(cr_year))
# year all
year_all <-
  hoaddata::jn_ind |>
  filter(issn_l != "0027-8424", !cr_year %in%  c("2017", "2023")) |>
  filter(issn_l %in% hoaddata::cc_articles$issn_l) |>
  distinct(issn_l, cr_year, jn_all) |> 
  group_by(cr_year) |> 
  summarise(n = sum(jn_all))

## Hybrid Journals
 n_hybrid_jns_all <- hoaddata::jn_ind |>
  filter(issn_l != "0027-8424", !cr_year %in%  c("2017", "2023")) |>
  distinct(issn_l) |>
   nrow()
 
## Hybrid Journals with at least one OA article
n_hybrid_jns <- hoaddata::cc_articles |>
  filter(issn_l != "0027-8424", !cr_year %in%  c("2017", "2023")) |>
  group_by(issn_l) |>
  summarise(oa_all = n_distinct(doi)) |> 
  nrow()

## Hybrid Journals with at least one OA article
n_ta_jns <- ta_oa_inst |>
  filter(ta_active == TRUE, !is.na(cc), issn_l != "0027-8424", !cr_year %in%  c("2017", "2023")) |>
  distinct(issn_l) |>
  nrow()

## Hybrid Journals with at least one TA-OA article
n_tas_with_oa <- ta_oa_inst |>
  filter(ta_active == TRUE, !is.na(cc), issn_l != "0027-8424", !cr_year %in%  c("2017", "2023")) |>
  distinct(ta_journal_portfolio) |>
  nrow()

## Open access volume
n_oa <-  hoaddata::cc_articles |>
  filter(issn_l != "0027-8424", !cr_year %in%  c("2017", "2023")) |>
  distinct(doi) |>
  nrow()

## Total
n_total <- hoaddata::jn_ind |>
  filter(issn_l != "0027-8424", !cr_year %in%  c("2017", "2023")) |>
  filter(issn_l %in% hoaddata::cc_articles$issn_l) |>
  distinct(issn_l, cr_year, jn_all) |>
  pull(jn_all) |>
  sum()

## TA total
n_tas_with_oa <- ta_oa_inst |>
  filter(ta_active == TRUE, !is.na(cc), issn_l != "0027-8424", !cr_year %in%  c("2017", "2023")) |>
  distinct(ta_journal_portfolio) |>
  nrow()

## TA article total
n_ta_oa <- ta_oa_inst |>
  filter(ta_active == TRUE, !is.na(cc), issn_l != "0027-8424", !cr_year %in%  c("2017", "2023")) |>
  distinct(doi) |>
  nrow()
```

Between 2018 and 2022, a total of `r n_hybrid_jns` out of `r n_hybrid_jns_all` hybrid journals in transformative agreements published at least one open access article under a Creative Commons license.
During this period, these hybrid journals provided open access to `r n_oa` out of `r n_total` articles, representing a five-year open access proportion of `r n_oa / n_total * 100`%.
Authors who could make use of transformative agreements at the time of publication contributed `r n_ta_oa` open access articles to the total.


```{r area_oa_over_the_years}
oa_ta_plot_df <- ta_df |>
  select(cr_year, ta_oa, no_ta) |>
  pivot_longer(cols = c(ta_oa, no_ta), names_to = "type", values_to = "value") |>
  filter(!cr_year %in%  c("2017", "2023")) |>
  inner_join(year_all)
  # plot
oa_ta_plot <- ggplot(oa_ta_plot_df, aes(cr_year, value / n, fill = type, group = type)) +
  geom_area() +
  scale_fill_manual(NULL, values = c(no_ta = "#b3b3b3a0", ta_oa = "#56B4E9"), labels = c(no_ta = "Other", ta_oa = "Agreement"),  guide = guide_legend(reverse = TRUE)) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.05)), labels = scales::percent_format(), limits = c(0, .15), breaks = c(0, .05, .1, .15)) +
  theme_minimal() +
  labs(y = "Hybrid OA Uptake", x = NULL) +
  theme(legend.position="top", legend.justification = "left", axis.text.x = element_blank())
```

```{r ta_share_plot}
ta_share_plot <- ta_df |>
  select(cr_year, ta_oa_prop, no_ta_prop) |>
  pivot_longer(cols = c(ta_oa_prop, no_ta_prop), names_to = "type", values_to = "value") |>
  filter(!cr_year %in%  c("2017", "2023"), type == "ta_oa_prop") |>
  ggplot(aes(gsub("^20", "'", as.character(cr_year)), value, group = 1)) +
  geom_bar(stat = "identity", fill = "#56B4E9") +
  scale_y_continuous(expand = expansion(mult = c(0, 0.05)), breaks = c(0, 0.25, 0.5),
                     labels = scales::percent) +
  theme_minimal() +
  labs(y = "Percentage\nof Hybrid OA\nvia Agreement", x = "Publication Year")
```

```{r journal_boxplot}
jn_hybrid_oa_growth <- hoaddata::cc_articles |>
  filter(issn_l != "0027-8424") |>
  group_by(cr_year, issn_l) |>
  summarise(oa_all = n_distinct(doi)) |>
  mutate(cr_year = factor(cr_year))
# year all
jn_year_all <-
  hoaddata::jn_ind |>
  filter(issn_l != "0027-8424") |>
  filter(issn_l %in% hoaddata::cc_articles$issn_l) |>
  distinct(issn_l, cr_year, jn_all) |> 
  group_by(cr_year, issn_l) |> 
  summarise(n = sum(jn_all))

oa_jn_box_df <- left_join(jn_hybrid_oa_growth, jn_year_all) |>
  mutate(prop = oa_all / n) |> 
  filter(!cr_year %in%  c("2017", "2023"))
oa_jn_box <- oa_jn_box_df |>
  ggplot(aes(gsub("^20", "'", as.character(cr_year)), prop)) +
  geom_boxplot(outlier.shape = NA) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.05)),
                     labels = scales::percent) +
  coord_cartesian(ylim = c(0,0.65)) +
  theme_minimal() +
  labs(y = "OA Uptake per Hybrid Journal", x = "Publication Year")
```

```{r}
ta_oa_growth <- ta_oa_inst |>
  filter(ta_active == TRUE, !is.na(cc), issn_l != "0027-8424") |>
  group_by(cr_year, issn_l) |>
  summarise(ta_oa = n_distinct(doi))
hybrid_oa_growth <- hoaddata::cc_articles |>
  filter(issn_l != "0027-8424") |>
  group_by(cr_year, issn_l) |>
  summarise(oa_all = n_distinct(doi))
ta_df <- left_join(hybrid_oa_growth, ta_oa_growth) |>
  mutate(across(everything(), .fns = ~replace_na(.,0))) |>
  mutate(no_ta = oa_all - ta_oa) |>
  mutate(cr_year = as.factor(cr_year))

summary_all <- inner_join(ta_df, jn_year_all, by = c("cr_year", "issn_l")) |>
  mutate(prop_ta_oa = ta_oa / n,
         prop_no_ta = no_ta / n) |>
  ungroup() |>
  distinct(cr_year, issn_l, prop_ta_oa, prop_no_ta) |>
  pivot_longer(cols = c(prop_ta_oa, prop_no_ta))
# plot
box_ta_prop <- summary_all |>
  filter(!cr_year %in%  c("2017", "2023")) |>
  # With at least one OA article
  filter(value > 0) |>
  ggplot(aes(gsub("^20", "'", as.character(cr_year)), value, fill = name)) +
  geom_boxplot(outlier.shape = NA, varwidth = TRUE) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.05)),
                     labels = scales::percent) +
  coord_cartesian(ylim = c(0,0.65)) +
  scale_fill_manual(NULL, values = c(prop_no_ta = "#b3b3b3a0", prop_ta_oa = "#56B4E9"), labels = c(prop_no_ta = "Other", prop_ta_oa = "Agreement"), guide = FALSE) +
  theme_minimal() +
  labs(y = NULL, x = "Publication Year")
  

```

```{r}
#| fig-cap: "Growth of open access in hybrid journals through transformative agreements between 2018 and 2022 per publication year. The blue areas represent open access through transformative agreements, as indicated by the affiliation with institutions that had a corresponding transformative agreement in place at the time of publication (according to the first author affiliation data in OpenAlex matched with journal and institutional data from cOAlition S Transformative Agreements Public Data). (A) Proportion of open access articles in hybrid journals per year (according to the total number of articles indexed in Crossref). (B) Percentage of hybrid open access via agreements per year. Boxplots show the proportion of open access articles by individual hybrid journals (C) and individual open access uptake rates by individual hybrid journals and open access funding (D) per publication year. Horizontal lines denote lower quartile, median, and upper quartile, with whiskers extending to ±1.5 of interquartile range. The individual outliers are not shown. Note that data on transformative agreements ending before June 2021 were not available at the time of this study."

library(patchwork)
layout = "
EEE
ACD
BCD"
wrap_plots(
  E = guide_area(),
  A = oa_ta_plot,
  B = ta_share_plot,
  C = oa_jn_box,
  D = box_ta_prop,
  design = layout,
  widths = c(0.4, 0.3, 0.3),
  heights = c(0.2, 1, .5),
  guides = "collect") &
  theme(
    legend.position = "top",
    legend.justification = "right",
    legend.direction = "horizontal",
    panel.grid.minor = element_blank(),
                panel.border = element_rect(color = "grey50", fill = NA),
    axis.title=element_text(size = 10)) &
    plot_annotation(tag_levels = 'A')
```

```{r}
# in text data
oa_by_year <- oa_ta_plot_df |>
  group_by(cr_year) |>
  summarise(oa = sum(value),
            articles = n) |>
  mutate(prop = oa / articles * 100) |>
  distinct()
```

Figure 2A shows a moderate growth in the proportion of open access articles in hybrid journals, comparing the overall open access uptake and the impact of transformative agreements on this trend.
Over the five-years period from 2018 to 2022, open access increased from `r oa_by_year |> filter(cr_year == "2018") |> pull(prop)`% (n = `r oa_by_year |> filter(cr_year == "2018") |> pull(oa)`) to `r oa_by_year |> filter(cr_year == "2022") |> pull(prop)`% (n = `r oa_by_year |> filter(cr_year == "2022") |> pull(oa)`). 
At the same time, the total article volume of the investigated journals grew from `r oa_by_year |> filter(cr_year == "2018") |> pull(articles)` in 2018 to `r oa_by_year |> filter(cr_year == "2022") |> pull(articles)` in 2022.

Figure 2B highlights that the majority of open access articles in hybrid journals were made available through transformative agreements in 2021 and 2022, contributing `r oa_ta_plot_df |> filter(cr_year == "2022", type == "ta_oa") |> pull(value) / oa_by_year |> filter(cr_year == "2022") |> pull(oa) * 100`% of the total open access article volume in 2022.
However, there was also a notable growth in open access provision through individual publication fees, which increased from `r oa_ta_plot_df |> filter(cr_year == "2018", type == "no_ta") |> mutate(prop = value / n * 100) |> pull(prop)`% (n = `r oa_ta_plot_df |> filter(cr_year == "2018", type == "no_ta") |> pull(value)`) in 2018 to  `r oa_ta_plot_df |> filter(cr_year == "2022", type == "no_ta") |> mutate(prop = value / n * 100) |> pull(prop)`% (n = `r oa_ta_plot_df |> filter(cr_year == "2022", type == "no_ta") |> pull(value)`). This suggests that publishers were able to gain equally from individual and institutional open access publishing options.

Figure 2C depicts the substantial variations among the hybrid journals included in transformative agreements in terms of open access uptake.
Although the median generally follows the trend shown in Figure 2A, the farther stretch of upper quartiles and whiskers over the years illustrates that an increasing number of journals published an above-average proportion of open access articles. 
In 2022, 25% of hybrid journals (n = `r oa_jn_box_df |> filter(cr_year == "2022") |> mutate(top25 = quantile(prop, .75)) |> filter(prop >= top25) |> nrow()`) had an open access uptake of `r oa_jn_box_df |> filter(cr_year == "2022") |> pull(prop) |> quantile(0.75)*100`%, and `r (oa_jn_box_df |> filter(cr_year == "2022", prop >= 0.5) |> nrow()) / length(unique(oa_jn_box_df$issn_l)) *100`% of journals (n = `r oa_jn_box_df |> filter(cr_year == "2022", prop >= 0.5) |> nrow()`) provided the majority of their articles under a Creative Commons license in the same year. 
These journals were, on average, smaller (M = `r oa_jn_box_df |> filter(cr_year == "2022", prop >= 0.5) |> pull(n) |> mean()`, SD = `r oa_jn_box_df |> filter(cr_year == "2022", prop >= 0.5) |> pull(n) |> sd()`) than those with an open access share below 50% (M = `r oa_jn_box_df |> filter(cr_year == "2022", prop <= 0.5) |> pull(n) |> mean()`, SD = `r oa_jn_box_df |> filter(cr_year == "2022", prop <= 0.5) |> pull(n) |> sd()`). 
Notable exception of large journals with an above-average open access proportion were Physical Review D, a high-energy physics journal covered by the SCOAP3 consortium that provided open access to `r oa_jn_box_df |> filter(cr_year == "2022", issn_l == "2470-0010") |> pull(oa_all)` out of `r oa_jn_box_df |> filter(cr_year == "2022", issn_l == "2470-0010") |> pull(n)` articles in 2022, Astronomy and Physics (`r oa_jn_box_df |> filter(cr_year == "2022", issn_l == "0004-6361") |> pull(oa_all)` out of `r oa_jn_box_df |> filter(cr_year == "2022", issn_l == "0004-6361") |> pull(n)` articles in 2022 were open access), which shifted to a subscribe to open business model for all accepted articles as of April 2022, the Journal of Fluid Mechanics (`r oa_jn_box_df |> filter(cr_year == "2022", issn_l == "0022-1120") |> pull(oa_all)` of `r oa_jn_box_df |> filter(cr_year == "2022", issn_l == "0022-1120") |> pull(n)` articles in 2022 were open access) and Bioinformatics (`r oa_jn_box_df |> filter(cr_year == "2022", issn_l == "1367-4803") |> pull(oa_all)` out of `r oa_jn_box_df |> filter(cr_year == "2022", issn_l == "1367-4803") |> pull(n)` articles in 2022 were open access), which flipped to full open access as of January 2023.

When comparing the impact of open access trough transformative agreements across journals, it shows that for many journals these agreements substantially contributed to the growth of open access over the years (Figure 2D). 
Examples of such journals include those with a scope on specific countries or regions, where also transformative agreements were implemented. 
For instance, in 2022, the Germany-based journals Zeitschrift für Erziehungswissenschaft and Zeitschrift für Politikwissenschaft, as well as the Scandinavian Political Studies adressing the Nordic countries, achieved an overall open access uptake of more than 90% just through transformative agreements. 
Despite the rise in transformative agreements, it is worth noting that other means of publishing open access in hybrid journals remained common. 
In total, `r summary_all |> filter(cr_year == "2022", value > 0, name == "prop_no_ta") |> nrow()` journals published open access articles from authors affiliated with institutions without transformative agreements in place, while `r summary_all |> filter(cr_year == "2022", value > 0, name == "prop_ta_oa") |> nrow()` journals published at least one open access article through a transformative agreement in the same year.

```{r}
jn_ind_jct <- hoaddata::jct_hybrid_jns |>
  inner_join(hoaddata::jn_ind, by = "issn_l") |>
  filter(issn_l != "0027-8424")

ta_active <- ta_oa_inst |>
  filter(issn_l != "0027-8424") |>
  filter(ta_active == TRUE, !is.na(cc)) |>
  group_by(cr_year, issn_l, esac_publisher) |>
  summarise(ta_oa = n_distinct(doi)) |>
  mutate(cr_year = as.character(cr_year))

jn_ind_jct_oa <- jn_ind_jct |>
  filter(!is.na(cc)) |>
  distinct(cr_year, issn_l, esac_publisher, cc, cc_total, jn_all) |>
  group_by(cr_year, issn_l, esac_publisher, jn_all) |>
  summarise(oa = sum(cc_total, na.rm = TRUE))

jn_ind_jct_all <- jn_ind_jct |>
  filter(issn_l %in% hoaddata::cc_articles$issn_l) |>
  distinct(cr_year, issn_l, esac_publisher, jn_all) 

publisher_df <- left_join(jn_ind_jct_all, jn_ind_jct_oa) |>
  mutate(cr_year = as.character(cr_year)) |>
  left_join(ta_active) |>
  mutate(across(everything(), .fns = ~replace_na(.,0))) |>
   mutate(no_ta = oa - ta_oa) |>
    mutate(
    publisher = case_when(
      esac_publisher == "Elsevier" ~ "Elsevier",
      esac_publisher == "Springer Nature" ~ "Springer Nature",
      esac_publisher == "Wiley" ~ "Wiley",
      is.character(esac_publisher) ~ "Other"
    )
  ) |>
  mutate(publisher =
           forcats::fct_relevel(
             publisher,
             c("Elsevier", "Springer Nature", "Wiley", "Other") 
           )) |>
  filter(!cr_year %in%  c("2017", "2023")) |>
  # Journal transfers across publishers grouped in publisher
  select(-esac_publisher) |>
  distinct()

## plot
publisher_plot_df <- publisher_df |> 
  pivot_longer(cols = c(ta_oa, no_ta), names_to = "type", values_to = "value")


#+ publisher ta shares
publisher_ta_prop <- publisher_plot_df |>
  filter(type == "ta_oa") |>
  group_by(cr_year, publisher) |>
  summarize(ta_oa = sum(value), oa = sum(oa), jn_all = sum(jn_all)) |>
  mutate(prop = ta_oa / oa) 


oa_publisher <- publisher_plot_df |>
  group_by(cr_year, publisher, type) |>
  summarise(value = sum(value))
all_publisher <- publisher_plot_df |>
  distinct(cr_year, issn_l, publisher, jn_all) |>
  group_by(cr_year, publisher) |>
  summarise(jn_all = sum(jn_all))

publisher_summary_df <- inner_join(oa_publisher, all_publisher) |>
  mutate(prop = value / jn_all)

publisher_prop <- ggplot(publisher_summary_df, aes(gsub("^20", "'", as.character(cr_year)), prop, fill = type, group = type)) +
  geom_area() +
  facet_wrap(~publisher, nrow = 1) +
  scale_fill_manual("Hybrid OA", values = c(no_ta = "#b3b3b3a0", ta_oa = "#56B4E9"), labels = c(no_ta = "Other", ta_oa = "Agreement"),  guide = guide_legend(reverse = TRUE)) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.05)), labels = scales::percent_format()) +
  theme_minimal() +
  labs(y = "Hybrid OA Uptake", x = NULL) +
  theme(legend.position="top", legend.justification = "left", axis.text.x = element_blank())

publisher_boxplot <- publisher_plot_df |>
  filter(value > 0) |>
  ggplot(aes(gsub("^20", "'", as.character(cr_year)), value / jn_all, fill = type)) +
  geom_boxplot(outlier.shape = NA, show.legend = FALSE, varwidth = TRUE) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.05)),
                     labels = scales::percent) +
  scale_fill_manual("Hybrid OA", values = c(no_ta = "#b3b3b3a0", ta_oa = "#56B4E9"), labels = c(no_ta = "Other", ta_oa = "Agreement")) +
  coord_cartesian(ylim = c(0,0.65)) +
  facet_wrap(~publisher, nrow = 1) +
  labs(y = "OA Uptake\nby Hybrid Journal", x = "Publication Year") +
  theme_minimal() +
    theme(strip.background = element_blank(),
    strip.text.x = element_blank()
    )

publisher_ta_prop_plot <- publisher_ta_prop |>
  ggplot(aes(gsub("^20", "'", as.character(cr_year)), prop)) +
  geom_bar(stat = "identity", fill = "#56B4E9") +
  scale_y_continuous(expand = expansion(mult = c(0, 0.05)), breaks = c(0, 0.25,0.5,0.75), limits = c(0, .75),
                     labels = scales::percent) +
  facet_wrap(~publisher, nrow = 1) +
  theme_minimal() +
  labs(y = "Percentage\nof Hybrid OA\nvia Agreement", x = NULL) +
  theme(
    axis.text.x = element_blank(),
    strip.background = element_blank(),
    strip.text.x = element_blank())
```

```{r publisher_summary_table}
publisher_article_total_table <- publisher_df |>
  group_by(publisher) |>
  summarise(oa_journals = n_distinct(issn_l),
            publisher_total = sum(jn_all)) |>
  mutate(
    jn_perc = oa_journals / sum(oa_journals),
    jn_cum = cumsum(oa_journals) / sum(oa_journals),
    perc = publisher_total / sum(publisher_total),
    cum = cumsum(publisher_total) / sum(publisher_total)
  ) |>
  relocate(publisher_total, .after = jn_cum)

publisher_article_oa_table <- publisher_summary_df |>
  group_by(publisher) |>
  summarise(oa_total = sum(value)) |>
  mutate(oa_perc = oa_total / sum(oa_total),
         oa_cum = cumsum(oa_total) / sum(oa_total))

publisher_article_ta_oa_table <- publisher_summary_df |>
  filter(type == "ta_oa") |>
  group_by(publisher) |>
  summarise(ta_oa_total = sum(value)) |>
  mutate(
    ta_oa_perc = ta_oa_total / sum(ta_oa_total),
    ta_oa_cum = cumsum(ta_oa_total) / sum(ta_oa_total)
  )

publisher_league_tbl <-
  inner_join(publisher_article_total_table, publisher_article_oa_table) |>
  inner_join(publisher_article_ta_oa_table)
```

```{r}
publisher_league_tbl |>
  mutate(across(contains(c("perc", "cum")), ~ paste(round(., 3) * 100))) |>
  kbl(
    caption ="Hybrid open access through transformative agreements market shares 2018-2022",
    col.names = c(
      "Publisher",
      "Total",
      "%",
      "C%",
      "Total",
      "%",
      "C%",
      "Total",
      "%",
      "C%",
      "Total",
      "%",
      "C%"
    ),
    format.args = list(big.mark = ',')
  ) %>%
  kable_classic() %>%
  add_header_above(c(
    " " = 1,
    "Hybrid journals" = 3,
    "Articles" = 3,
    "OA articles" = 3,
    "TA OA articles" = 3
  ))
```

Analysing hybrid open access across publishers between 2018 and 2022 reveals a large market concentration. 
Although `r length(unique(jn_ind_jct_all$esac_publisher))` publishers offered transformative agreements, the big three commercial publishers Elsevier, Springer Nature, and Wiley, accounted for `r publisher_league_tbl  |> filter(publisher == "Wiley") |> pull(jn_cum) * 100`% of total article volume published (see Table ).
Together, they published `r publisher_league_tbl  |> filter(publisher != "Other") |> pull(oa_total) |> sum()` open access articles, representing `r publisher_league_tbl  |> filter(publisher == "Wiley") |> pull(oa_cum) * 100`% of the open access articles in hybrid journals.
Elsevier, Springer Nature, and Wiley made `r publisher_league_tbl  |> filter(publisher != "Other") |> pull(ta_oa_total) |> sum()` articles open access in hybrid journals through transformative agreements, resulting in an even larger market share of `r publisher_league_tbl  |> filter(publisher == "Wiley") |> pull(ta_oa_cum)*100`%.
However, there are differences among the three large publishers. Although Elsevier published the largest volume of articles (n = `r publisher_league_tbl  |> filter(publisher == "Elsevier") |> pull(publisher_total)`, `r publisher_league_tbl  |> filter(publisher == "Elsevier") |> pull(cum)*100`%), it published a relatively low number of open access articles, including those that can be associated with transformative agreements.
In contrast, Springer Nature and Wiley provided open access to a larger proportion of their articles (`r (publisher_league_tbl  |> filter(publisher == "Springer Nature") |> pull(oa_total)) / (publisher_league_tbl  |> filter(publisher == "Springer Nature") |> pull(publisher_total)) * 100`% of Springer Nature articles and `r (publisher_league_tbl  |> filter(publisher == "Wiley") |> pull(oa_total)) / (publisher_league_tbl  |> filter(publisher == "Wiley") |> pull(publisher_total)) * 100`% of Wiley articles were open acccess), leading to higher open access market shares (`r publisher_league_tbl  |> filter(publisher == "Springer Nature") |> pull(oa_perc)*100`% Springer Nature resp. `r publisher_league_tbl  |> filter(publisher == "Springer Nature") |> pull(oa_perc)*100`% Wiley).
This  difference between Elsevier on the one hand and Springer Nature and Wiley on the other can  be attributed to transformative agreements, as the latter made the majority of their open access articles available through such deals (Springer Nature `r ((publisher_league_tbl  |> filter(publisher == "Springer Nature") |> pull(ta_oa_total)) / (publisher_league_tbl  |> filter(publisher == "Springer Nature") |> pull(oa_total)) * 100)`% resp. Wiley `r (publisher_league_tbl  |> filter(publisher == "Wiley") |> pull(ta_oa_total)) / (publisher_league_tbl  |> filter(publisher == "Wiley") |> pull(oa_total)) * 100`%).


Figure 2 takes a closer look into the growth of hybrid open access across publishers by year with a focus on open articles enabled by transformative agreements. 
Although all publishers show a general long-term trend towards transformative agreements, Figure 2A and 2B indicate that, in particular, the open access share of Wiley's hybrid journal portfolio quadrupled from `r publisher_summary_df |> filter(cr_year == "2018", publisher == "Wiley") |> pull(prop) |> sum()*100`% in 2018 to `r publisher_summary_df |> filter(cr_year == "2022", publisher == "Wiley") |> pull(prop) |> sum()*100`% in 2022 as a result of these agreements. 
In comparison, the proportion of open access articles Elsevier hybrid journals grew from from `r publisher_summary_df |> filter(cr_year == "2018", publisher == "Elsevier") |> pull(prop) |> sum()*100`% in 2018 to `r publisher_summary_df |> filter(cr_year == "2022", publisher == "Elsevier") |> pull(prop) |> sum()*100`% in 2022. 
Unlike the other two publishers, Springer Nature started early to offer selected consortia open access agreements for its hybrid journal portfolio ("Springer Compact").
Data about the Springer compact agreements with large consortia including Max Planck Society, the Swedish Bibsam consortium qnd the Finish	FinELib consortium that ended before start of the trnasofrmative agreement data collection in June 2021.
 `r publisher_summary_df |> filter(cr_year == "2018", publisher == "Springer Nature") |> pull(prop) |> sum()*100`% Springer Nature excelled in 2018, but gained a slower growth than Wiley, resulting into `r publisher_summary_df |> filter(cr_year == "2022", publisher == "Springer Nature") |> pull(prop) |> sum()*100`% of all articles being open access in Springer Nature hybrid journals in 2022.



published the less than half of its `r publisher_summary_df |> filter(cr_year == "2022", publisher == "Elsevier") |> pull(value) |> sum()` open access articles in 2022 through transformative agreements (`r publisher_summary_df |> filter(cr_year == "2022", publisher == "Elsevier", type == "ta_oa") |> pull(value) / publisher_summary_df |> filter(cr_year == "2022", publisher == "Elsevier") |> pull(value) |> sum()*100`%).  while `r publisher_summary_df |> filter(cr_year == "2022", publisher == "Springer Nature", type == "ta_oa") |> pull(value) / publisher_summary_df |> filter(cr_year == "2022", publisher == "Springer Nature") |> pull(value) |> sum()*100`% of Springer Nature's grew slower in the last year compared to Wiley, and  `r publisher_summary_df |> filter(cr_year == "2022", publisher == "Wiley", type == "ta_oa") |> pull(value) / publisher_summary_df |> filter(cr_year == "2022", publisher == "Wiley") |> pull(value) |> sum()*100`% of Wiley's open access articles could be linked to transformative agreements in 2022. Overall, Wiley published the largest number of open access articles from transformative agreements in 2022 (n = `r publisher_summary_df |> filter(cr_year == "2022", publisher == "Wiley") |> pull(value) |> sum()`), followed by Springer Nature (`r publisher_summary_df |> filter(cr_year == "2022", publisher == "Wiley") |> pull(value) |> sum()`) and Elsevier (`r publisher_summary_df |> filter(cr_year == "2022", publisher == "Wiley") |> pull(value) |> sum()`). This varying focus resulted into an different overall open access proportion in 2022 -- only 10% of all articles in Elsevier hybrid journals were open access in 2022, compared to published in springer nautre and xx in Wiley hybrid journals. 

The increasing impact of open access through transofmrative agreements can be also obsevered at the journal-level (Figure 2). While no substantial differences between open access enabled through transformative agreements can be seen across Elsevier journals, an growing number of hybrid journals published by Springer Nature and Wiley increased their open access share through these agreements. In 2022, 25% of Springer Nature hybrid journal portfolio comprised journals with an open access proportion of   percent, while Wiley. 

again a glanc eon the jorunals indicates that these ar emore smaller journals focus on a researchers affitialted with a socienty or a language springe rnautre german-speaking didactic journasl whereas in wiley british society tend to 
across all more smaller like 
smaller and 
of journals majority and springe rnature majority oabove 50 % . prominent example BJC 


```{r, fig.asp=1.1}
library(patchwork)
design = "E
A
B
C"
wrap_plots(
  E = guide_area(),
  A = publisher_prop,
  B = publisher_ta_prop_plot,
  C = publisher_boxplot,
  design = design,
  guides = "collect",
  heights = c(0, 0.8, 0.3, 0.8)
) &
  theme(
    legend.position = "top",
    legend.justification = "right",
    legend.direction = "horizontal",
    panel.grid.minor = element_blank(),
                panel.border = element_rect(color = "grey50", fill = NA),
    axis.title=element_text(size = 10)) &
    plot_annotation(tag_levels = 'A') 
```
